---
title: 'Peer-graded Assignment: Course Project 1'
author: "ChristopherLadner"
date: "January 19, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the 'quantified self' movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

* Dataset: Activity monitoring data [52K]

The variables included in this dataset are:

* **steps:** Number of steps taking in a 5-minute interval (missing values are coded as NA)
* **date:** The date on which the measurement was taken in YYYY-MM-DD format
* **interval:** Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

Review criterialess 
Repo

Valid GitHub URL
At least one commit beyond the original fork
Valid SHA-1
SHA-1 corresponds to a specific commit
Commit containing full submission

Code for reading in the dataset and/or processing the data
Histogram of the total number of steps taken each day
Mean and median number of steps taken each day
Time series plot of the average number of steps taken
The 5-minute interval that, on average, contains the maximum number of steps
Code to describe and show a strategy for imputing missing data
Histogram of the total number of steps taken each day after missing values are imputed
Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

## Assignment

This assignment will be described in multiple parts. You will need to
write a report that answers the questions detailed below. Ultimately,
you will need to complete the entire assignment in a **single R
markdown** document that can be processed by **knitr** and be
transformed into an HTML file.

Throughout your report make sure you always include the code that you
used to generate the output you present. When writing code chunks in
the R markdown document, always use `echo = TRUE` so that someone else
will be able to read the code. **This assignment will be evaluated via
peer assessment so it is essential that your peer evaluators be able
to review the code for your analysis**.

For the plotting aspects of this assignment, feel free to use any
plotting system in R (i.e., base, lattice, ggplot2)

Fork/clone the [GitHub repository created for this
assignment](http://github.com/rdpeng/RepData_PeerAssessment1). You
will submit this assignment by pushing your completed files into your
forked repository on GitHub. The assignment submission will consist of
the URL to your GitHub repository and the SHA-1 commit ID for your
repository state.

NOTE: The GitHub repository also contains the dataset for the
assignment so you do not have to download the data separately.


### Loading and preprocessing the data

Show any code that is needed to

Question #1. Load the data (i.e. **read.csv()**)

```{r load}
if (!file.exists("data")) {
  dir.create("data")
}
## Download the data and load into the activity Dataframe 
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./Data/ActivityMonitoringDataset.zip")
unzip(zipfile="./Data/ActivityMonitoringDataset.zip", exdir = "./Data")
activity = read.csv("./data/activity.csv")  # read csv file
```

Question #2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r view and mutate activity}
summary(activity)
str(activity)
head(activity)

library(dplyr)
## Rearrange columns in a logical time series
activity <- select(activity, date, interval, steps)
```

```{r activity}
head(activity)
```


### What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.
```{r remove NAs}
activity.noNAs <- na.omit(activity)
```

Calculate the total number of steps taken per day
```{r summary}
steps.per.day <- activity.noNAs %>% group_by(date) %>% 
  summarize(total.steps.taken=sum(steps))
head(steps.per.day)
```

Question #1. Make a histogram of the total number of steps taken each day

```{r histogram}
library(ggplot2)
ggplot(data=steps.per.day, aes(total.steps.taken)) + 
      geom_histogram()
```

Question #2. Calculate and report the **mean** and **median** total number of steps taken per day

```{r mean and median}
mean(steps.per.day$total.steps.taken)
median(steps.per.day$total.steps.taken)
```

### What is the average daily activity pattern?
Q1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r time series plot}
library(lattice)
steps.by.interval <- activity.noNAs %>% group_by(interval) %>% 
  summarize(average.steps = mean(steps))
xyplot(average.steps ~ interval, steps.by.interval, type="l")
```

Q2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max interval}
filter(steps.by.interval, average.steps == max(average.steps))
```

### Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Q1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r number of NAs}
sum(is.na(activity$steps) == TRUE)
```

Q2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r strategy to replace NAs}
activity.imputed <- merge(activity, steps.by.interval, by="interval")
head(activity.imputed)

##Function to replace step's NAs
for (i in 1:nrow(activity.imputed)) {
  if (is.na(activity.imputed[i,3]== TRUE)) {
      activity.imputed$steps[i]=
      activity.imputed$average.steps[i]
  }
}
head(activity.imputed)

##Remove the average.steps column - no longer needed
 activity.imputed$average.steps=NULL
 head(activity.imputed)
 
##Rearrange the dataframe
activity.imputed <- select(activity.imputed, date, interval, steps)
activity.imputed <- arrange(activity.imputed, date)
head(activity.imputed)

##Confirm all NAs were imputed
sum(is.na(activity.imputed$steps)==TRUE)
```
Q4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r Histogram of imputed total steps taken}
imputed.steps.per.day <- activity.imputed %>% group_by(date) %>% 
  summarize(total.steps.taken=sum(steps))
head(imputed.steps.per.day)

## Create the histogram
library(ggplot2)
ggplot(data=imputed.steps.per.day, aes(total.steps.taken)) +       geom_histogram()

## Calculate and report the mean and median total number of steps ## taken per day.
mean(imputed.steps.per.day$total.steps.taken)
median(imputed.steps.per.day$total.steps.taken)
```
Do these values differ from the estimates from the first part of ## the assignment? the mean is the same while the median has increased

What is the impact of imputing missing data on the estimates of ## the total daily number of steps? the median changed but the distribution is the same

### Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Q1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r create weekday variable}
## Use the weekdays() function to create the DayOffWeek variable
activity.imputed <- mutate(activity.imputed,DayOfWeek=
  weekdays(as.Date(activity.imputed$date)))
head(activity.imputed)

## Create function to translate DayOfWeek to Weekend or Weekday
for (i in 1:nrow(activity.imputed)) {
  if (activity.imputed[i,"DayOfWeek"] == "Saturday" || activity.imputed[i,"DayOfWeek"] == "Sunday") {
      activity.imputed[i,"TypeOfDay"] = "Weekend"
 }else {
    activity.imputed[i,"TypeOfDay"] = "Weekday"
    }
}
head(activity.imputed)

## Create summary dataframe for ave steps by interval/TypeOfDay
steps.by.interval.act.imputed <- activity.imputed %>% 
  group_by(TypeOfDay, interval) %>%   
  summarize(average.steps = mean(steps))

## View he dataframe
head(steps.by.interval.act.imputed)
```

Q2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r create panel plot}
ggplot(data = steps.by.interval.act.imputed, 
  aes(x=interval, y=average.steps)) +
  geom_line() +
  facet_grid(TypeOfDay ~ .) +
  xlab("5 minute interval") + 
  ylab("average number of steps taken") +
  ggtitle("Activity Patterns Between Weekdays and Weekends")
```


